---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("cbsodataR")
install.packages("rnaturalearth")
install.packages("sf")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(cbsodataR)
require(rnaturalearth)
require(sf)
```

# Title Page


Joris Vanniyasingam 2812507
Jard van Duin 2883878
Roy de Wit 2862758
Tristan Dijkstra 2864813
David Lot 2784019

Group 05 06

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

The Dutch housing market has seen a substantial increases in housing prices over the past 2 decades. As housing becomes less affordable, it becomes clear that income is not increasing at a rate which can keep up with the rising costs of housing. This report analyses how housing costs and income change, looking at the 12 separate dutch provinces. By doing this, it makes it possible to analyse if the changes in housing prices and income differ per province, making it possible to see if a province with higher income also has higher housing prices compared to provinces with a lower income. This report could then be a helpful tool for deciding in which province it is most economically efficient to buy a house in. 

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

#data_verkoopprijs <- cbs_get_data("83625NED")
#data_income <- cbs_get_data("86163NED")

#saveRDS(data_verkoopprijs, "data/verkoopprijs.rds")
#saveRDS(data_income, "data/income.rds")

data_verkoopprijs <- readRDS("data/verkoopprijs.rds")
view(data_verkoopprijs)
head(data_verkoopprijs)

data_income <- readRDS("data/income.rds")
view(data_income)
head(data_income)

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

```{r data_cleaning}
# het filteren van de code zodat alleen de provincies overblijven
library(dplyr)

data_income_cleaned <- data_income %>%
  filter(startsWith(RegioS, "PV")) %>%
  select("BedrijfstakkenBranchesSBI2008", "RegioS", "Perioden", "GemiddeldPersoonlijkInkomen_2") %>%
  group_by(RegioS, Perioden) %>%
  summarise(Gemiddeld_Persoonlijk_inkomen = mean(GemiddeldPersoonlijkInkomen_2, na.rm = TRUE),            )

table(data_income$RegioS)

```

```{r data_cleaning}
library(dplyr)

data_verkoopprijs_cleaned <- data_verkoopprijs %>%
  filter(startsWith(RegioS, "PV")) %>%
  filter(Perioden >= "2007JJ00", Perioden <= "2024JJ00")
```

```{r data_merging}
merged_dataset <- inner_join(data_income_cleaned, data_verkoopprijs_cleaned, by = c("RegioS", "Perioden"))

merged_dataset <- merged_dataset %>%
  mutate(Perioden = substr(Perioden, 1,4)) %>%
  mutate(Perioden = as.numeric(Perioden))

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
# variable aanmaken verschil % inkomen
merged_dataset <- merged_dataset %>%
  mutate(Percentage_verandering_inkomen = (Gemiddeld_Persoonlijk_inkomen - lag(Gemiddeld_Persoonlijk_inkomen)) / lag(Gemiddeld_Persoonlijk_inkomen) * 100)
```

Variable 2

```{r gen_var2}
# variable aanmaken verschil % verkoopprijs
merged_dataset <- merged_dataset %>%
  mutate(Percentage_verandering_verkoopprijs = (GemiddeldeVerkoopprijs_1 - lag(GemiddeldeVerkoopprijs_1)) / lag(GemiddeldeVerkoopprijs_1) * 100)
```

Variable 3
```{r gen_var3}
#deze is optioneel voor als we nog iets extra's willen
```

## 3.3 Visualize temporal variation

```{r}
mean_procentuele_verandering_inkomen <- aggregate(
  Percentage_verandering_inkomen ~ Perioden,
  data = merged_dataset,
  mean,
  na.rm = TRUE
)

mean_procentuele_verandering_verkoopprijs <- aggregate(
  Percentage_verandering_verkoopprijs ~  Perioden,
  data = merged_dataset,
  mean,
  na.rm = TRUE
)

plot1 <-ggplot() +
  geom_line(data = mean_procentuele_verandering_inkomen,
            aes(x = Perioden, y = Percentage_verandering_inkomen, color = "Inkomen")) +
  geom_line(data = mean_procentuele_verandering_verkoopprijs,
            aes(x = Perioden, y = Percentage_verandering_verkoopprijs, color = "Verkoopprijs")) +
  labs(
    title = "Gemiddelde procentuele verandering",
    x = "Jaren",
    y = "Procentuele verandering"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(2008, 2024, by = 1),
  )

plot(plot1)
  


```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(sf)

provincies_kaart <- st_read("https://service.pdok.nl/cbs/gebiedsindelingen/2023/wfs/v1_0?service=WFS&version=2.0.0&request=GetFeature&typeName=provincie_gegeneraliseerd&outputFormat=application/json")

data_kaart <- merged_dataset %>%
  mutate(RegioS = trimws(RegioS)) %>%
  filter(Perioden == "2021")

provincies_map <- provincies_kaart %>%
  left_join(data_kaart, by = c("statcode" = "RegioS"))

colnames(provincies_map)

ggplot(provincies_map) +
  geom_sf(aes(fill = GemiddeldeVerkoopprijs_1)) +
  scale_fill_gradient(
     low = "#ffcccc",
     high = "#8b0000",
    name = "Gemiddelde\nVerkoopprijs (€)",
    labels = scales::comma
  ) +
  labs(
    title = "Gemiddelde Verkoopprijs Koopwoningen per Provincie",
    subtitle = "2021",
    caption = "Bron: CBS"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )


```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
provincies_kaart <- st_read("https://service.pdok.nl/cbs/gebiedsindelingen/2023/wfs/v1_0?service=WFS&version=2.0.0&request=GetFeature&typeName=provincie_gegeneraliseerd&outputFormat=application/json")

data_kaart <- merged_dataset %>%
  mutate(RegioS = trimws(RegioS)) %>%
  filter(Perioden == "2021") %>%
  mutate(Regio = case_when(
    RegioS %in% c("PV20", "PV21", "PV22") ~ "Noord",
    RegioS %in% c("PV23", "PV24", "PV25") ~ "Oost",
    RegioS %in% c("PV27", "PV28", "PV26") ~ "West",
    RegioS %in% c("PV29", "PV30", "PV31") ~ "Zuid",
    TRUE ~ "Onbekend"
  ))

regio_data <- data_kaart %>%
  group_by(Regio) %>%
  summarise(GemPrijs = mean(GemiddeldeVerkoopprijs_1, na.rm = TRUE))

provincies_map <- provincies_kaart %>%
  left_join(data_kaart, by = c("statcode" = "RegioS")) %>%
  group_by(Regio) %>%
  summarise(GemPrijs = mean(GemiddeldeVerkoopprijs_1, na.rm = TRUE),
            geometry = st_union(geometry))

ggplot(provincies_map) +
  geom_sf(aes(fill = GemPrijs)) +
  scale_fill_gradient(
    low = "#cce5ff",
    high = "#003366",
    name = "Gemiddelde\nVerkoopprijs (€)",
    labels = scales::comma
  ) +
  geom_sf_label(aes(label = Regio), size = 4, color = "red") +
  labs(
    title = "Gemiddelde Verkoopprijs per Regio",
    caption = "Bron: CBS"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
mean_verkoopprijs <- aggregate(
  GemiddeldeVerkoopprijs_1 ~  Perioden,
  data = merged_dataset,
  mean,
  na.rm = TRUE
)

plot2 <-ggplot() +
  geom_line(data = mean_verkoopprijs,
            aes(x = Perioden, y = GemiddeldeVerkoopprijs_1, color = "Verkoopprijs")) +
  labs(
    title = "Verkoopprijs over de jaren",
    x = "Jaren",
    y = "Verkoopprijs woningen"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(2008, 2024, by = 4),
  ) +
  geom_vline(aes(xintercept=2020, color="Covid-19"), linetype="dashed")

plot(plot2)
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/1066121414/EBE_programming_05_06

## 5.2 Reference list

Use APA referencing throughout your document.

Centraal Bureau voor de Statistiek. (2025, 19 november). Zelfstandigen; inkomen, vermogen, regio (indeling 2025). Centraal Bureau Voor de Statistiek. https://www.cbs.nl/nl-nl/cijfers/detail/86163NED#shortTableDescription 
